Synthesizing Symmetric Lenses

### 2019-02-28 papers/SymmetricLenses/main.tex
turned off drafts
Comments and polishing

\newif\ifdraft\draftfalse  % set true to show comments
% \newif\ifdraft\draftfalse  % set true to show comments
\newif\ifanon\anontrue    % set true to suppress names, etc.
\newif\ifappendices\appendicesfalse

@@

{\em Lenses} are programs that can be run both ``front to back'' and ``back
to front,'' allowing updates to both input and output data to be transferred
in both directions.
Since their introduction by Foster \ETAL, lenses have been extensively
studied, extended, and applied.  Recent work has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.

\hspace*{1em}We show how to extend this synthesis algorithm to a much wider class of
lenses, which we call {\em simple symmetric lenses}, including not only all
of the bijective lenses and the more widely used ``asymmetric''
lenses, but also a rich subset of the full-blown
``symmetric lenses'' proposed by Hofmann \ETAL  
Intuitively, simple symmetric lenses allow some information to be present
on one side but not the other and vice versa.
Simple symmetric lenses are of independent theoretical interest, being
the largest class of symmetric lenses that do not rely on persistent
internal state.  
Synthesizing simple symmetric lenses is substantially more challenging than
synthesizing bijective lenses: Since some of the information on each side
can be ``disconnected'' from the other side, there will, in general, be {\em
  many} lenses that agree with a given example. To guide the search process,
we use {\em stochastic regular expressions} and ideas from information
theory to estimate the amount of information propagated by a candidate
lens, generally prefering lenses that propagate more information, 
as well as user annotations identifying parts of the source and
target data structures as either {\em irrelevant} or {\em essential}.

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language. We evaluate
its performance on 48 benchmark examples drawn from Flash Fill,
Augeus, the bidirectional programming literature, and electronic
file format synchronization tasks. Our implementation can synthesize all these
lenses in under 30 seconds.
\end{abstract}

@@

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing ``round tripping'' guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.
Domain-specific languages for writing lenses can facilitate generation of
principled data transformations.  For example, Boomerang~\cite{boomerang} is a
language for expressing bidirectional string transformations.

\textit{Synthesizing} lenses can make the task even easier. Indeed, lens
languages, like Boomerang, are excellent targets for search-based synthesis, since their
specialized, sub-Turing-complete syntax and expressive type systems drastically
limit the space of possible programs.
%
The recent Optician~\cite{optician} enhancement to Boomerang has
demonstrated that it is feasible to synthesize quite complex examples, albeit
only for a restricted class of lenses---so-called bijective lenses.
Given source and target formats plus a small number of examples of
corresponding data, Optician's synthesis algorithm is guaranteed to find a
bijective lens matching the examples, if one exists. Indeed, it can often find
an appropriate lens in less than a second with no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection, it is fairly likely to be the intended one.

However, the set of real-world use cases for bijective lenses is also
somewhat limited.  A more common situation is where two real-world data
formats share just {\em some} of their information content.  For instance,
two related database tables might share four out of five columns, but each
might have a different fifth column.  Or, one ad-hoc system configuration
file might include some format-specific metadata, such as a date or a
reference number, whereas the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic, 13 benchmarks in
total.

>>>>

\newif\ifdraft\drafttrue  % set true to show comments
% \newif\ifdraft\draftfalse  % set true to show comments
\newif\ifanon\anontrue    % set true to suppress names, etc.
\newif\ifappendices\appendicesfalse

@@

{\em Lenses} are programs that can be run both ``front to back'' and ``back
to front,'' allowing updates to either their source or their target data to be
transferred in both directions. 
Since their introduction by Foster \ETAL, lenses have been extensively
studied, extended, and applied.  Recent work has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.

\hspace*{1em}We extend this synthesis algorithm to a much broader class of
lenses, called {\em simple symmetric lenses}, including all
bijective lenses, all of the popular category of ``asymmetric''
lenses, and a rich subset of the more powerful
``symmetric lenses'' proposed by Hofmann \ETAL  
Intuitively, simple symmetric lenses allow some information to be present
on one side but not the other and vice versa.
They are of independent theoretical interest, being
the largest class of symmetric lenses that do not rely on persistent
internal state.

Synthesizing simple symmetric lenses is substantially more challenging than
synthesizing bijective lenses: Since some of the information on each side
can be ``disconnected'' from the other side, there will, in general, be {\em
  many} lenses that agree with a given example. To guide the search process,
we use {\em stochastic regular expressions} and ideas from information
theory to estimate the amount of information propagated by a candidate
lens, generally prefering lenses that propagate more information, 
as well as user annotations marking parts of the source and
target data structures as either {\em irrelevant} or {\em essential}.

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language. We evaluate
its performance on 48 benchmark examples drawn from Flash Fill,
Augeus, the bidirectional programming literature, and electronic
file format synchronization tasks. Our implementation can synthesize all these
lenses in under 30 seconds.  \bcp{All of them together take 30 seconds?  Or
  the maximum time it takes for one is 30 seconds?}
\end{abstract}

@@

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing round-tripping guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.
Domain-specific languages for writing lenses can facilitate generation of
principled data transformations.  For example, Boomerang~\cite{boomerang} is a
language for expressing bidirectional string transformations.

\textit{Synthesizing} lenses can make the task even easier. Indeed, lens
languages like Boomerang are excellent targets for search-based synthesis,
since their 
specialized, sub-Turing-complete syntax and their expressive type systems
drastically limit the space of possible programs.
%
The recent Optician synthesizer for Boomerang~\cite{optician} has
demonstrated the feasibility of synthesizing quite complex examples, albeit
only for a restricted class of lenses---so-called bijective lenses.
Given source and target formats plus a small number of examples of
corresponding data, Optician's synthesis algorithm is guaranteed to find a
bijective lens matching the examples, if one exists. Indeed, it can often find
an appropriate lens in less than a second with no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm finds any bijection, it is fairly likely to be
the intended one.

However, the set of real-world use cases for bijective lenses, where two
data formats contain different arrangements of precisely the same
information, is  
limited.  A more common situation is where two data
formats share just {\em some} of their information content.  For instance,
two related database tables might share four out of five columns, but each
might have a different fifth column.  Or one ad-hoc system-configuration
file might include some format-specific metadata, such as a date or a
reference number, while the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic, 13 benchmarks in
total.  \bcp{I don't understand this.  We say that the non-bijective
  situation is ``more common,'' but we support this by saying that 13 out of
48 benchmarks (that we chose!) are non-bijective.  Or am I misreading??}
